An Algorithm for Constructing Monotone \\ Quintic Interpolating Splines

### 2019-12-15 Paper_[2019-11]_HPC/paper/paper.tex
Nearly done with HPC paper. Incorporated Dr. Watson revisions. Running third experiment more extensively for final figures.

@@

%***************************************************************************
% AUTHOR: AUTHOR NAMES GO HERE
% FORMAT AUTHORS NAMES Like: Author1, Author2 and Author3 (last names)
%
%		You need to change the author listing below!
%               Please list ALL authors using last name only, separate by a comma except
%               for the last author, separate with "and"

@@

% AUTHOR: Enter the title, all letters in upper case
\title{An Algorithm for Constructing Monotone Piecewise \\ Quintic Hermite Interpolating Polynomials}

% AUTHOR: Enter the authors of the article, see end of the example document for further examples
\author{
Thomas C. H. Lux\\
Layne T. Watson\\
Tyler H. Chang\\ [12pt]
Department of Computer Science\\
Virginia Polytechnic Institute and State University $\,$\\
2000 Torgersen Hall, Blacksburg, VA, USA\\
\{tchlux,ltwatson,thchang\}@vt.edu\\
\and
Li Xu\\%
Yueyao Wang\\%
Yili Hong\\ [12pt]
Department of Statistics\\
$\,$ Virginia Polytechnic Institute and State University\\
213 Hutcheson Hall, Blacksburg, VA, USA\\
yilihong@vt.edu\\
}

@@

An algorithm for computing monotone piecewise quintic interpolating
polynomials is proposed. Algebraic constraints for enforcing
monotonicity are provided that align with quintic monotonicity
theory. The algorithm is implemented, tested, and applied to several
sample problems to demonstrate the improved accuracy of piecewise
quintic monotone interpolation compared to the existing piecewise
cubic solution.

\textbf{Keywords:} monotone, spline, Hermite interpolation, quintic polynomial.

@@

Many domains of science rely on smooth approximations to real-valued functions over a closed interval. Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate non parametric approximations in statistics \cite{knott2012interpolating}. While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.

It is often the case that the true underlying function or phenomenon being modeled has known properties e.g., convexity, positivity, various levels of continuity, or monotonicity. Given a reasonable amount of data, it quickly becomes difficult to achieve desirable properties in a single polynomial function. In general, the maintenance of function properties through interpolation / regression is referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}. The specific shapes this work will achieve in approximations are monotonicity and two levels of continuity. These properties are chiefly important to the approximation of cumulative distribution functions and subsequently the effective generation of random numbers from a specified distribution.

In statistics especially, the construction of a monotone interpolating spline that is continuous in second derivative is meaningfully useful \cite{ramsay1988monotone}. A function with these properties could approximate random variables to a high level of accuracy with relatively few intervals. A continuously twice differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a property many standard parametric distributions maintain.

The currently available software for monotone piecewise polynomial interpolation includes quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and (with limited application) quartic \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases. In addition, a statistical method for bootstrapping the construction of an arbitrarily smooth monotone fit exists \cite{leitenstorfer2006generalized}, but the method does not take advantage of known analytic properties related to quintic polynomials. Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, however this theory has not yet been used to construct a piecewise quintic monotone spline interpolation routine. Recent work suggests that the lack of quintic software may be due to a general unawareness of the theory \shortcite{xie2018semiparametric}.

\begin{figure}
  \centering
  \includegraphics[width=.9\textwidth]{spline-demonstration}
  \caption{Example polynomials that interpolate function values at the ends of the interval $[0,1]$. The first only interpolates the function values $f(0) = 1$ and $f(1) = 0$, making it the order two polynomial $f(x) = 1 - x$. For the second plot $f(x) = 8x^3 - 12x^2 + 4x + 1/2$, which is order four and interpolates the values $f(0) = 1/2$, $f'(0) = 4$, $f(1) = 1/2$, $f'(1) = 4$. Finally the third plot shows the order six polynomial $f(x) = - 64x^5 + 160x^4 - 140x^3 + 50x^2 - 7x + 1$ interpolating the function values $f(0) = 1$, $f'(0) = -7$, $f''(0) = 100$, $f(1) = 0$, $f'(1) = -7$, $f''(1) = -100$. Notice that interpolating the same fixed number of function values at each endpoint will always result in an even order interpolating polynomial.
  }\label{fig:spline-demonstration}
\end{figure}

The importance of piecewise quintic interpolation over lower order approximations can be simply demonstrated. In general, the order of a polynomial determines the number of function values it can interpolate, and the growth rate of error away from the interpolated function values. As demonstrated in Figure \ref{fig:spline-demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two. The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on generic splines.

@@

The major contribution of this work is an algorithm for constructing piecewise quintic Hermite interpolating polynomials that utilizes existing quintic monotonicity theory. The remainder of this paper is structured as follows: ...

@@

The current state of the art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ and $\{f(k_i) : i = 1,2,\ldots,n\}$ be a given set of data values at the partition points for a monotone function $f$, meaning $f(k_i) \leq f(k_{i+1})$ for $i = 1, \ldots, n-1$ or $f(k_i) \geq f(k_{i+1})$ for $i = 1, \ldots, n-1$. Let $\hat f$ be a piecewise cubic spline defined in each sub-interval $I_i = [k_i, k_{i+1}]$ by

\begin{align*}
h_i =& \ k_{i+1} - k_{i} \\
u(t) =& \ 3t^2 - 2t^3 \\
p(t) =& \ t^3 - t^2 \\
\hat f(x) =& \ f(k_i)\ u\big((k_{i+1} - x) / h_i\big) + f(k_{i+1})\ u\big((x - k_i) / h_i\big) \\
& - \hat f^{\ \prime}(k_i)\ p\big((k_{i+1}-x)/h_i\big) + \hat f^{\ \prime}(k_{i+1})\ p\big((x-k_i)/h_i \big).
\end{align*}

Notice that a trivially monotone spline results when $\hat f^{\ \prime}(k_i) = 0$, for $i = 1, \ldots, n$. However, such a spline has too many \textit{wiggles} for most applications. Fritsch and Carlson show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals. Consider the terms $\alpha = \big(\hat f^{\ \prime}(k_i) (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\hat f^{\ \prime}(k_{i+1}) (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$, now monotonicity of a cubic polynomial over a sub-interval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.

\begin{figure}
  \centering
  \includegraphics[width=0.50\textwidth]{demo_projection}
  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone.}\label{fig:projection}
\end{figure}

The actual region of monotonicity for a cubic polynomial is larger, but projection of $(\alpha, \beta)$ into one of these regions ensures that monotonicity will be achieved and not violated for neighboring regions. The user must decide which region is most appropriate for the projections based on the application, Fritsch and Carlson recommend using region 2.

@@

%% - show picture of how the cubic is computed
%% - show the algorithm for computing a cubic (and that it is trivial)
%% - quintic interpolation is more accurate when data is available, or higher order information can be inferred

@@

The following section is composed of three algorithms that together are used to construct a piecewise quintic monotone interpolating spline. Section \ref{is_monotone} regards checking monotonicity, Section \ref{make_monotone} presents a procedure for enforcing monotonicity on a polynomial piece, and Section \ref{monotone_spline} combines the previous two algorithms into a routine for enforcing monotonicity over a quintic spline.

Similar to the cubic case, let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ and $\{f(k_i) : i = 1,2,\ldots,n\}$ be a given set of values at the partition points for a monotone function $f$, meaning $f(k_i) \leq f(k_{i+1})$ or $f(k_i) \geq f(k_{i+1})$ for $i = 1, \ldots, n-1$.

In pseudo code, a function is used that is referred to as \texttt{line\_search($g,$ $a,$ $b$)}, where $a,$ $b$ $\in S$ for $S$ closed under convex combination, $g: S \rightarrow \{0,1\}$ is a binary function, and $g(b) = 1$. The \texttt{line\_search} is performed by Binary search in the presented framework, when $g(a) = 1$, $a$ is returned, otherwise the smallest $c \in [0,1]$ such that $g\big(a(1-c) + c b\big) = 1$ is returned.

In the case that first or second derivative information is not available at time of approximation, it is estimated with the central divided difference of the function values. The final quintic spline is represented with an exact rational arithmetic Newton Polynomial. % and evaluated with a B-spline basis.

\vspace{10pt}

@@

\subsection{Verifying Monotonicity of a Quintic Polynomial}
\label{is_monotone}
Let $f$ be a quintic polynomial over a closed interval $[x_0, x_1]$ $\subset \mathbb{R}$. Now $f$ is uniquely defined by the evaluation tuples $\big(x_0,$ $f(x_0),$ $f'(x_0),$ $f''(x_0)\big)$ and $\big(x_1,$ $f(x_1),$ $f'(x_1),$ $f''(x_1)\big).$ Assume without loss of generality that $f(x_0) < f(x_1),$ where the case of monotonic decreasing $f$ would consider the negated the function values. The following algorithm will determine whether or not $f$ is monotone increasing on the interval $[x_0, x_1].$

\vspace{10pt}%
\hrule%
\vspace{3pt}%
\noindent\textbf{\textit{Algorithm 1:}} \texttt{is\_monotone$\big (x_0$,$x_1$,$f \big)$}%
\vspace{3pt}%
\hrule%

\begin{itemize}
  \itemsep0pt
  \parskip0pt

\item[1:] \texttt{if $\big(f(x_0) = f(x_1)\big)$ return $\big( Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) = 0 \big)$}
\item[2:] \texttt{if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$ return FALSE}
  \begin{itemize}
    \item[] \textit{This can be seen clearly from the fact that $f$ is analytic; there will exist some nonempty interval about $x_0$ or $x_1$ for which $Df$ is negative.}
  \end{itemize}

\item[2:] \texttt{if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$}
\item[3:] $\quad w = x_0 - x_1$
\item[4:] $\quad g(y) = \big({-4} Df(x_1) / w \big) < \big( ({-f}(x_1)/w - 24 y - 32 Df(x_1)) / 5w \big)$
\item[5:] $\quad$ \texttt{line\_search}$(g, 0, Df(x_0))$
\item[6:] $A = Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
\item[7:] $B = Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
  \begin{itemize}
    \item[] \textit{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \cite{ulrich1994positivity}, first defined after equation 18.}
  \end{itemize}
\item[8:] $\gamma_0 = 4 \frac{Df(x_0)}{Df(x_1)} (B/A)^{3/4}$
\item[9:] $\gamma_1 = \frac{x_1 - x_0}{Df(x_1)} (B/A)^{3/4}$
\item[4:] $\alpha_0 = 4 (B/A)^{1/4}$
\item[5:] $\alpha_1 = -\frac{x_1 - x_0}{Df(x_1)} (B/A)^{1/4}$
\item[6:] $\beta_0 = 30 - \frac{12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
\item[7:] $\beta_1 = \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $
  \begin{itemize}
    \item[] \textit{The $\gamma,$ $\alpha,$ and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in \cite{ulrich1994positivity} (equation 16). These terms with subscripts $0$ and $1$ give the computation of $\alpha,$ $\beta,$ and $\gamma$ the form seen in lines $10$-$12$ below.}
  \end{itemize}
\item[10:] $\gamma = \gamma_0 + \gamma_1 D^2f(x_0)$
\item[11:] $\alpha = \alpha_0 + \alpha_1 D^2f(x_1)$
\item[12:] $\beta = \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$
\item[13:] \texttt{if $(\beta \leq 6)$ then return $\alpha > - (\beta + 2) / 2$}
\item[14:] \texttt{else return $\gamma > -2 \sqrt{\beta - 2}$ }

\end{itemize}
\hrule
\vspace{10pt}
%% ----------------------------------------------------------------------

%% ----------------------------------------------------------------------
Given the same initial conditions there are special circumstances which allow for the usage of simpler monotonicity conditions. In this case, consider when the quintic function has either $f'(x_0) = 0$ or $f'(x_1) = 0.$ This reduces the problem of verifying monotonicity to one of cubics established by \cite{schmidt1988positivity}.

\vspace{10pt}%
\hrule%
\vspace{3pt}%
\noindent\textbf{\textit{Algorithm 1b:}} \texttt{is\_monotone\_simplified}%
\vspace{3pt}%
\hrule%
\begin{itemize}
  \itemsep0pt
  \parskip0pt

>>>>

%% Use a custom `algorithm` command for this paper.
\input{lux-algorithm}

@@

%***************************************************************************
% AUTHOR: Thomas C.H. Lux et al.
% FORMAT AUTHORS NAMES Like: Author1, Author2 and Author3 (last names)
%
%		You need to change the author listing below!
%               Please list ALL authors using last name only, separate by a comma except
%               for the last author, separate with "and"

@@

% AUTHOR: Enter the title, all letters in upper case
\title{An Algorithm for Constructing Monotone \\ Quintic Interpolating Splines}

% AUTHOR: Enter the authors of the article, see end of the example document for further examples
\author{
Thomas C. H. Lux\\
Layne T. Watson\\
Tyler H. Chang\\ [12pt]
Departments of Computer Science, Mathematics,\\
and Aerospace and Ocean Engineering\\
Virginia Polytechnic Institute \& State University\\
2000 Torgersen Hall, Blacksburg, VA, USA\\
\{tchlux,ltw,thchang\}@vt.edu\\
\and
Li Xu\\%
Yueyao Wang\\%
Yili Hong\\ [12pt]
Department of Statistics\\
Virginia Polytechnic Institute \& State University\\
213 Hutcheson Hall, Blacksburg, VA, USA\\
yilihong@vt.edu\\
}

@@

An algorithm for computing monotone order six piecewise polynomial
interpolants is proposed. Algebraic constraints for enforcing
monotonicity are provided that align with quintic monotonicity
theory. The algorithm is implemented, tested, and applied to several
sample problems to demonstrate the improved accuracy of monotone
quintic spline interpolants compared to that for monotone cubic spline
interpolants.

\textbf{Keywords:} monotone, quintic spline, Hermite interpolation, sixth order polynomial.

@@

Many domains of science rely on smooth approximations to real-valued functions over a closed interval. Piecewise polynomial functions (splines) provide the smooth approximations for animation in graphics \cite{herman2015techniques,quint2003scalable}, aesthetic structural support in architecture \cite{brennan2019measure}, efficient aerodynamic surfaces in automotive and aerospace engineering \cite{brennan2019measure}, prolonged effective operation of electric motors \shortcite{berglund2009planning}, and accurate nonparametric approximations in statistics \cite{knott2012interpolating}. While polynomial interpolants or regressors apply broadly, splines are often a good choice because they can approximate globally complex functions while minimizing the local complexity of an approximation.

It is often the case that the true underlying function or phenomenon being modeled has known properties e.g., convexity, positivity, various levels of continuity, or monotonicity. Given a reasonable amount of data, it quickly becomes difficult to achieve desirable properties in a single polynomial function. In general, the maintenance of function properties through interpolation/regression is referred to as \textit{shape preserving} \cite{fritsch1980monotone,gregory1985shape}. The specific shapes this work will achieve in approximations are monotonicity and $C^2$ continuity. These properties are chiefly important to the approximation of cumulative distribution functions and subsequently the effective generation of random numbers from a specified distribution.

In statistics especially, the construction of a monotone interpolating spline that is $C^2$ continuous is meaningfully useful \cite{ramsay1988monotone}. A function with these properties could approximate a cumulative distribution function to a high level of accuracy with relatively few intervals. A twice continuously differentiable approximation to a cumulative distribution function (CDF) would also produce a corresponding probability density function (PDF) that is continuously differentiable, which is a property many standard parametric distributions maintain.

The currently available software for monotone piecewise polynomial interpolation includes quadratic \cite{he1998monotone}, cubic \cite{fritsch1980monotone}, and (with limited application) quartic \cite{wang2004rational,piah2011improved,yao2018unconditionally} cases. In addition, a statistical method for bootstrapping the construction of an arbitrarily smooth monotone fit exists \cite{leitenstorfer2006generalized}, but the method does not take advantage of known analytic properties related to quintic polynomials. Theory has been provided for the quintic case \cite{ulrich1994positivity,hess1994positive}, however this theory has not yet been used to construct a monotone quintic spline interpolation routine. Recent work suggests that the lack of quintic software may be due to a general unawareness of the theory \shortcite{xie2018semiparametric}.

\begin{figure}
  \centering
  \includegraphics[width=.9\textwidth]{spline_demonstration}
  \caption{Example polynomials that interpolate function values at the ends of the interval $[0,1]$. The first only interpolates the function values $g_2(0) = 1$ and $g_2(1) = 0$, making it the order two polynomial $g_2(x) = 1 - x$. For the second plot $g_4(x) = 8x^3 - 12x^2 + 4x + 1/2$, which is order four and interpolates the values $g_4(0) = 1/2$, $g_4'(0) = 4$, $g_4(1) = 1/2$, $g_4'(1) = 4$. Finally the third plot shows the order six polynomial $g_6(x) = - 64x^5 + 160x^4 - 140x^3 + 50x^2 - 7x + 1$ interpolating the function values $g_6(0) = 1$, $g_6'(0) = -7$, $g_6''(0) = 100$, $g_6(1) = 0$, $g_6'(1) = -7$, $g_6''(1) = -100$. Notice that interpolating the same fixed number of function values at each endpoint will always result in an even order interpolating polynomial.
  }\label{fig:spline_demonstration}
\end{figure}

The importance of piecewise quintic interpolation over lower order approximations can be simply demonstrated. In general, the order of a polynomial determines the number of function values it can interpolate, and the growth rate of error away from the interpolated function values. As demonstrated in Figure \ref{fig:spline_demonstration}, it can be seen that matching a value at either end of the interval requires an order two (linear) approximation and each additional derivative at the ends of the interval raises the necessary polynomial order by two. The body of this work is composed of a novel algorithm for enforcing monotonicity on quintic polynomial pieces, then extending that solution to work on quintic splines.

@@

The major contribution of this work is an algorithm for constructing monotone quintic interpolating splines that utilizes existing quintic monotonicity theory. The remainder of this paper is structured as follows: {\bf ?????}

@@

The current state-of-the-art monotone interpolating spline with a mathematical software implementation is piecewise cubic, continuously differentiable, and was first proposed in \cite{fritsch1980monotone} then expanded upon in \cite{carlson1985monotone}. Let $\pi: x_0 = k_1 < k_2 < \cdots < k_n = x_1$ be a partition of the interval $[x_0,x_1]$. Let $f: \mathbb{R} \rightarrow \mathbb{R}$ be $C^2$, and $\bigl\{f(k_i)\bigr\}_{i=1}^n$ and $\bigl\{\Delta_i\bigr\}_{i=1}^n$ be given sets of function and derivative values at the partition points for a monotone function $f$. Either $f(k_i) \leq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\ge0$, $i=1$, $\ldots$, $n$, or $f(k_i) \geq f(k_{i+1})$, $i=1$, $\ldots$, $n-1$, and $\Delta_i\le0$, $i=1$, $\ldots$, $n$. Let $\hat f$ be a piecewise cubic polynomial defined in each subinterval $I_i = [k_i, k_{i+1}]$ by

\begin{align*}
h_i &= k_{i+1} - k_{i}, \quad
u(t) = 3t^2 - 2t^3, \quad
p(t) = t^3 - t^2,\\
\hat f(x) &= f(k_i) u\big((k_{i+1} - x) / h_i\big) +
f(k_{i+1}) u\big((x - k_i) / h_i\big) \\
&{} - h_i\Delta_i p\big((k_{i+1}-x)/h_i\big) + h_i\Delta_{i+1}
p\big((x-k_i)/h_i\big).
\end{align*}

Notice that a trivially monotone spline results when $\Delta_i = 0$, for $i = 1$, $\ldots$, $n$. However, such a spline has too many \textit{wiggles} for most applications. \citeN{carlson1985monotone} show that simple conditions on the derivative values can guarantee monotonicity, and that these conditions can be enforced in a way that ensures modifications on one interval will not break the monotonicity of cubic polynomials over any neighboring intervals. Consider the terms $\alpha = \big(\Delta_i (k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$ and $\beta = \big(\Delta_{i+1}(k_{i+1}-k_i)\big) / \big(f(k_{i+1}) - f(k_i)\big)$. If $f(k_{i+1}) - f(k_i)=0$, take $\Delta_i = \Delta_{i+1} =0$ and $\alpha =\beta =1$. In general, monotonicity of a cubic polynomial over a subinterval can be maintained by ensuring that $\alpha$ and $\beta$ reside in any of the regions depicted in Figure \ref{fig:projection}.

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{cubic_projection_demonstration}
  \caption{These are the feasible regions of monotonicity for cubic splines and the projections that make a cubic polynomial piece monotone. The regions themselves are numbered 1--4 corresponding to their original description in Fritsch and Carlson (1980). One point is projected onto each region as a demonstration.}\label{fig:projection}
\end{figure}

The actual region of monotonicity for a cubic polynomial is larger, but projection of $(\alpha, \beta)$ into one of these regions ensures that monotonicity will be achieved and not violated for neighboring regions. The user must decide which region is most appropriate for the projections based on the application, Fritsch and Carlson recommend using Region 2.

@@

@@

The following section is composed of three algorithms that together are used to construct a monotone quintic spline interpolant. Without loss of generality, the algorithms will only consider the monotone increasing (nondecreasing) case. The monotone decreasing (nonincreasing) case is handled similarly. Algorithm \ref{alg:check-monotone} checks monotonicity,
Algorithm \ref{alg:make-monotone} enforces monotonicity on an order six
polynomial piece of the quintic spline, and Algorithm
\ref{alg:monotone-spline} uses the previous two algorithms to enforce
monotonicity for the entire quintic spline.

In pseudocode, a function \texttt{binary\_search($g,$ $a,$ $b$)} is used, where $a,b\in S\subset \mathbb{R}^p$ for convex $S$, $g: S \rightarrow \{$\texttt{FALSE}, \texttt{TRUE}$\}$ is a right continuous Boolean function, $g(b) = \hbox{\texttt{TRUE}}$, and for $\mu \in [0,1]$, $g\big((1-\mu)a + \mu b\big) =\hbox{\texttt{TRUE}} \Longrightarrow g\big((1-\nu)a + \nu b\big) =\hbox{\texttt{TRUE}}$ for $\mu \le \nu\le1$. The search returns $\big((1-c)a + c b\big)$ for the smallest $c \in [0,1]$ such that $g\big((1-c)a + c b\big) = \hbox{\texttt{TRUE}}$.

These algorithms make use of first and second derivatives of the approximated function. In the case that the first and second derivative information is not provided along with function values, the derivatives are estimated with finite differences of the function values. The final quintic spline is represented as a piecewise polynomial using the Newton form for each polynomial piece, and ultimately converted to a $B$-spline representation for evaluation.

\algorithm
    {is\_monotone$\big(x_0, x_1, f \big)$}
    {where $x_0$, $x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial. Returns \texttt{TRUE} if $f$ is monotone increasing on $[x_0,x_1]$.}
    {\label{alg:check-monotone}
  \item if $\big(f(x_0) = f(x_1)\big)$ and not $\big( 0 = Df(x_0) = Df(x_1) = D^2f(x_0) = D^2f(x_1) \big)$
    \indented{return FALSE}
  \item if $\big(Df(x_0) < 0$ or $Df(x_1) < 0\big)$
    \indented{return FALSE}
    \comment{The necessity of these first two steps follows directly from the fact that $f$ is $C^2$. The next case that will be covered is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$. Step \ref{alg1-alpha} checks for the necessary condition that $\alpha \ge 0$, Step \ref{alg1-beta} checks $\beta \geq \alpha$, and Step \ref{alg1-gamma} checks $\gamma \geq \delta$, all from \citeN{schmidt1988positivity}. If all necessary conditions are met, then the order six piece is monotone and Step \ref{alg1-simplified-true} concludes this check.}
  \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
  \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
  \item \indented{if $\big(D^2f(x_1) > {-4}Df(x_1) / w \big)$ return FALSE}
    \label{alg1-alpha}
  \item \indented{if $\big(D^2f(x_1) < (3w D^2f(x_0) - 24 Df(x_0) - 32
    Df(x_1) + 60v/w) / (5w) \big)$ return FALSE} \label{alg1-beta}
  \item \indented{if $\big(D^2f(x_0) < 3 Df(x_0) / w \big)$ return FALSE}
    \label{alg1-gamma}
  \item \indented{return TRUE} \label{alg1-simplified-true}
    \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
  \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \comment{The variables $A$ and $B$ correspond directly to the theoretical foundation for positive quartic polynomials established in \citeN{ulrich1994positivity}, first defined after Equation (18).}
  \item $\displaystyle \gamma_0 := 4 \frac{Df(x_0)}{Df(x_1)} (B/A)^{3/4}$
    \\ $\displaystyle \gamma_1 := \frac{x_1 - x_0}{Df(x_1)} (B/A)^{3/4}$
    \\ $\alpha_0 := 4 (B/A)^{1/4}$
    \\ $\displaystyle \alpha_1 := -\frac{x_1 - x_0}{Df(x_1)} (B/A)^{1/4}$
    \\ $\displaystyle \beta_0 := 30 - \frac{12 \big(Df(x_0) + Df(x_1)\big) (x_1 - x_0)}{\big(f(x_1) - f(x_0)\big) \sqrt{A}\sqrt{B}}$
    \\ $\displaystyle \beta_1 := \frac{-3 (x_1 - x_0)^2}{2 \big(f(x_1) - f(x_0)\big) \sqrt{A} \sqrt{B}} $ \label{alg1-linearized-abg}
    \comment{The $\gamma,$ $\alpha,$ and $\beta$ terms with subscripts $0$ and $1$ are algebraic reductions of the simplified conditions for satisfying Theorem 2 in Equation (16) of \citeN{ulrich1994positivity}. These terms with subscripts $0$ and $1$ make the computation of $\alpha,$ $\beta,$ and $\gamma$ functions of the second derivative terms, as seen in Step \ref{alg1-abg} below.}
  \item $\gamma := \gamma_0 + \gamma_1 D^2f(x_0)$
    \\ $\alpha := \alpha_0 + \alpha_1 D^2f(x_1)$
    \\ $\beta := \beta_0 + \beta_1 \big(D^2f(x_0) - D^2f(x_1)\big)$
    \label{alg1-abg}
  \item if $(\beta \leq 6)$ then return $\big(\alpha > - (\beta + 2) / 2\big)$
  \item else return $\big(\gamma > -2 \sqrt{\beta - 2}\big)$ \label{alg1-last}
}
%% ----------------------------------------------------------------------

The reason for structuring the $\alpha$, $\beta$, and $\gamma$ computations in terms of the second derivative of $f$ (seen in Step \ref{alg1-abg} of Algorithm \ref{alg:check-monotone}) will become more apparent later. The next problem to consider is that of making a nonmonotone order six polynomial piece into a monotone one by modifying its first and second derivative values at the ends of an interval. Note that the actual value of the function at the ends of the interval is not modified, as the resulting polynomial needs to interpolate.

@@

\algorithm
    {make\_monotone$\big (x_0$,$x_1$,$f \big)$}
    {where $x_0,\ x_1 \in \mathbb{R}$, $x_0 < x_1$, and $f$ is an order six polynomial. Returns monotone $f$.}
    {\label{alg:make-monotone}
    \item if $\big(f(x_1) - f(x_0) = 0\big)$
      \indented{$Df(x_0) := Df(x_1) := D^2f(x_0) := D^2f(x_1) := 0$}
      \indented{return $f$}
    \item $Df(x_0) :=\ $median$\displaystyle \bigg(0,\ Df(x_0),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0} \bigg)$
      \\ $Df(x_1) :=\ $median$\displaystyle \bigg(0,\ Df(x_1),\ 14 \frac{f(x_1) - f(x_0)}{x_1 - x_0}\bigg)$ \label{alg2-init-df}
      \comment{The selection of values in Step \ref{alg2-init-df} for $Df(x_0)$ and $Df(x_1)$ is suggested by \citeN{ulrich1994positivity} and originally from \citeN{huynh1993accurate}. These rules quickly enforce upper and lower bounds on derivative values to ensure quintic monotonicity is obtainable.}
      \comment{The next case that will be covered is in accordance with a simplified condition for quintic monotonicity that reduces to one of cubic positivity studied in \citeN{schmidt1988positivity}, where $\alpha$, $\beta$, $\gamma$, and $\delta$ are defined in terms of values and derivatives of $f$ at $x_0$ and $x_1$. Steps \ref{alg2-dx-conditions} and \ref{alg2-rescale-dx} ensure that the following bounds on $D^2f(x_0)$ are compatible. Step \ref{alg2-ddx0-upper} is the final preparatory step, ensuring that the conditions on $D^2f(x_1)$ are compatible. To conclude the simplified case, Step \ref{alg2-gamma} first enforces $\gamma \geq \delta$, then Steps \ref{alg2-alpha} and \ref{alg2-beta} enforce $\alpha > 0$ and $\beta \geq \alpha$ respectively. Combined, these three conditions derived from Equation (2.18) in \citeN{schmidt1988positivity} guarantee monotonicity of this polynomial piece.}
    \item if $\big(Df(x_0) = 0$ or $Df(x_1) = 0\big)$
    \item \indented{$w := x_0 - x_1$\\$v := f(x_0) - f(x_1)$}
    \item \indented{if $\big(5 Df(x_0) + 4Df(x_1) > 20v / w)$} \label{alg2-dx-conditions}
    \item \indented{\indented{$Df(x_0) := Df(x_0) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}}
      \indented{\indented{$Df(x_1) := Df(x_1) \text{ max}\displaystyle \bigg(0, \frac{20v}{w \big(5 Df(x_0) + 4 Df(x_1)\big)}\bigg) $}} \label{alg2-rescale-dx}
    \item \indented{$D^2f(x_0) :=\ $min$\displaystyle \bigg(D^2f(x_0), \frac{4 (2 Df(x_0)+Df(x_1))+20 v/w}{w} \bigg)$} \label{alg2-ddx0-upper}
    \item  \indented{$D^2f(x_0) :=\ $max$\big(D^2f(x_0), 3 Df(x_0) / w \big)$} \label{alg2-gamma}
    \item \indented{$D^2f(x_1) :=\ $min$\big(D^2f(x_1), {-4}Df(x_1) / w \big)$} \label{alg2-alpha}
    \item  \indented{$D^2f(x_1) :=\ $max$\displaystyle \bigg(D^2f(x_1), \frac{3w D^2f(x_0) - 24 Df(x_0) - 32 Df(x_1) + 60v/w}{5w} \bigg)$} \label{alg2-beta}
    \item
      \indented{return $f$}
      \comment{The following code considers the remaining case where $Df(x_0) \neq 0$ and $Df(x_1) \neq 0$.}
    \item $\displaystyle A := Df(x_0)\frac{x_1 - x_0}{f(x_1) - f(x_0)}$
      \\ $\displaystyle B := Df(x_1) \frac{x_1 - x_0}{f(x_1) - f(x_0)}$
    \item if $\big($max$(A,B) > 6\big)$
      \indented{$Df(x_0) := 6 Df(x_0)\ /\ $max$(A,B)$}
      \indented{$Df(x_1) := 6 Df(x_1)\ /\ $max$(A,B)$}
      \comment{This box bound ensures that $(A,B)$ remains within a viable region of monotonicity (satisfying Theorem 4, seen in Fig. 6 of \citeN{ulrich1994positivity}).}
    \item $\eta := \big(D^2f(x_0), D^2f(x_1)\big)$
      \\ $\displaystyle \eta_0 := \bigg({-\frac{\sqrt{A}}{4}}\big(7\sqrt{A} + 3\sqrt{B}\big),\ \frac{\sqrt{B}}{4}\big(3\sqrt{A} + 7\sqrt{B}\big)  \bigg)$
    \item $D^2f(x_0),\ D^2f(x_1) := $ binary\_search$\big($is\_monotone$(x_0,x_1,f)$, $\eta,$ $\eta_0\big)$
